\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-26 paper.tex
edits for S7. Remove strong assertions about identifiability

The graph visualisations in Fig.~\ref{fig-simplification} have
three novel features.
Firstly, edge weights (the thickness of the lines joining
nodes) correspond to the length of the inheritance intervals
they are annotated with. This allows us to distinguish
edges that persist across many local trees from those that are
less influential (contrast the edge
$(\noderef{g}, \noderef{h})$
with $(\noderef{g}, \noderef{i})$
in Fig.~\ref{fig-simplification}A).
% FIXME this needs some work
Secondly, node colours denote the number of parents that they
have in the graph, allowing us to easily see roots (those
with zero parents), recombinants (those with two parents)
and more complex situations arising from simplification (see below).
Thirdly, the shading intensity of a node denotes the ``coalescent span'', the
fraction of the node's span (the length of genome in which it
is reachable from the samples in the local trees)
over which it is coalescent (has more than one child). Nodes which
are never locally unary therefore have a coalescent span of 100\%,
whereas nodes in which ancestral material never coalesces have a
coalescent span of 0\% (such nodes will be extremely difficult to
identify from the distribution of shared mutations in the samples).
% This is useful why? I can't come up with a quick way of saying
% this. In some senses it's a measure of identifiability?
% JK: Hmm, not sure we want to jump into inferability here?
All these features will impact what kind of an ARG we can
hope to infer from sampled genomes.

Returning to the main topic of this section,
Fig.~\ref{fig-simplification}A is the direct output of a Wright-Fisher
simulation in which we retain all nodes involved in recombination
or common ancestry events. This is the true history, and contains
a very high level of detail, some of which may be considered
redundant (or, from another perspective, unobservable).
In Fig.~\ref{fig-simplification}A the local trees (right)
contain many unary nodes, and as we successively simplify,
the local trees contain fewer
unary nodes (Fig.~\ref{fig-simplification}B,C) until
we reach Fig.~\ref{fig-simplification}D, where the local trees
have no unary nodes.
A node is unary in a local tree covering the genome interval $I$
if genetic inheritance passes through that ARG node,
but no coalescence occurs in the interval $I$.
The distinction between the ``common ancestry'' of two or more genomes
in an ancestral genome and the ``coalescence'' which may or may
not occur in the local trees is
important~\citep{hudson1983testing,kelleher2016efficient}.
Consider \noderef{e} in Fig.~\ref{fig-simplification}A,
for example. We can see from the graph that it is a common
ancestor of samples \noderef{a} and \noderef{b}, but
it does not correspond to any coalescence in the
local trees to the left of position $44$, and is therefore
unary in these three trees.

The first level of simplification that we can perform is to remove
parts of the graph topology that are invisible to the samples.
An example of such topology is a
``diamond''~\citep{rasmussen2014genome}
in which the two parent nodes of a recombination immediately
join again into a common ancestor (e.g.~\noderef{j}, \noderef{l}, \noderef{m}
and \noderef{n} in Fig.~\ref{fig-simplification}A).
Unless we are specifically
interested in the recombination event or these ancestral genomes,
there is no information in this topology and the diamond can be
replaced by a single edge. More generally, any
subgraph that is singly-connected in both the leafward and
rootward direction (a ``super-diamond'') is non-identifiable and can be
replaced by one edge. This definition includes the case
of a node that has one inbound and one outbound edge, such as
nodes \noderef{f} and \noderef{h}.
Fig.~\ref{fig-simplification}B shows the result of this type of
graph topology simplification.

Simplifying away diamonds will remove many unary nodes from the
local trees, but there can still be nodes that are unary in all
of the local trees. In particular, a node can represent a recombinant
with multiple parents in the graph but only a single child (e.g.\ node \noderef{n}
in Fig.~\ref{fig-simplification}B), or can represent a common ancestor with
multiple children in the graph but in which no coalescence takes place
in the local trees
(node \noderef{r} in Fig.~\ref{fig-simplification}B).
% point out that this is why Hudson refers to CA nodes, not coalescence nodes
Such nodes are not singly connected in the graph, but are nevertheless unary in
all of the local trees.
The operation to remove them
therefore requires knowledge not just of the graph topology but also of the
ancestral material associated with the edges.
As we see in Fig.~\ref{fig-simplification}C,
removal of recombinant nodes can produce graph nodes with
more than two parents (e.g.~node \noderef{e}); and likewise, removal of
common ancestor but non-coalescent nodes can produce graph nodes with
more than two children (e.g.~node \noderef{s}). These represent the
merged
\emph{effects} of multiple evolutionary events in a single node (genome), and the
ARG no longer contains the intermediate genomes representing those events.

@@

The ARGs in Fig.~\ref{fig-simplification} contain different
numbers of local trees ($6$, $5$, $5$ and $4$ respectively for A through
D). When we move from A to B the local trees
for the intervals $[44,61)$ and $[61,87)$ are merged because
the only differences between them are their paths through
nodes \noderef{l} and \noderef{m}. These nodes that participated
in the diamond are removed from the ARG, and we have lost
all information about the corresponding recombination at
position 61. Other nodes (e.g.\ \noderef{o} and \noderef{p})
have also been removed but these represent the \emph{parents}
of recombinants. The recombinant nodes themselves
(e.g. \noderef{n}) are still present, and represent precise
information about the time, genomic location and tree
branches involved
in the recombination event.

@@

% The point of these examples is to emphasise that
% information about recombination in an ARG is not ``all
% or nothing'': there is a gradation of detail
% about recombination, in terms of genomic location, timing,
% and the ancestral lineages involved.
% While there has been a recent tendency to emphasise the
% importance of ``full ARGs''
% \citep[e.g.][]{deng2021distribution,brandt2021evaluation,
% rasmussen2022espalier}
% which contain complete information about all recombination
% events, it is reasonable to question the value of
% such precise information and whether it is inferrable.
% [SOME MORE STUFF]

% Ancestral recombination graphs are often characterised
% as consisting of a complete record of the genetic history
% of a sample. For example,
% \cite{rasmussen2014genome} state that an ``ARG provides a record of all
% coalescence and recombination events since the divergence of the sequences
% under study'' and
% according to \cite{deng2021distribution} an ARG
% ``provides all the information about the genealogical history of a sample,
% including the locations of recombination events.'' It is worth
% questioning, however, whether such comprehensive information
% will always be available, and in particular whether our
% representation of genealogical history should \emph{require}
% such precision. Simulations will usually generate complete
% information about the history of a sample, of course, but
% we may not have sufficient information to infer such detail from sampled genomes.

% % The full ancestral recombination graph (ARG) is a structure that encodes all
% % coalescence and recombination events resulting from the stochastic process of
% % the coalescent with recombination.
% \citet{brandt2021evaluation} define a ``full'' ARG as
% ``a structure that encodes all
% coalescence and recombination events resulting from the stochastic process of
% the coalescent with recombination'' (thus going against the general
% historial trend discussed in the XXX section).
% \citet{rasmussen2022espalier} have a less restrictive definition, in
% which a ``full'' ARG is one that contains complete information about
% recombination events, without necessarily being tied to a particular
% stochastic process.
% While [something positive], it is important to be clear on
% what the overall goal of such precise inferences are,
% what the limitations on when they can be made are,
% and their downstream utility in applications.

% % What is the goal of having fully precise ARGs?
% % 1) studying the actual recombination events;
% % 2) evaluating the likelihoods
% The main reasons for wanting precise estimates of all recombination
% events are to either compute a likelihood of the inferred ARG
% under a statistical model, or to study the recombinants in
% detail~\citep{rasmussen2022espalier}.
% % Explain that we're really interested in the details of recombination
% % itself with a small number of samples.
% In the latter case, we might
% have a small number of samples~\citep{guo2022recombination}.

% Beyond cases in which visual inspection of the trees and the
% effects of recombination is feasible, the main application
% of having completely precise estimates of recombination events is
% to evaluate the likelihood of the estimate under some
% statistical model. The only statistical model available
% is the coalescent with recombination
% and its approximations~\citep{mcvean2005approximating,marjoram2006fast}.
% In this case, the utility of a ``full`` ARG for some dataset
% is limited by how well the dataset fits the assumptions
% of the model.

% % JK: This is very rough, just jotting down the basic ideas,
% % will need substantial revision.
% It is not clear that such complete information will always
% (or indeed ever) be available from inferences~\cite{shipilina2023origin}.
%  We may be
% able to precisely identify the location and timing
% of recent recombinations (particularly if detailed pedigree
% information is available) but beyond this, the information
% required simply isn't present. When sampling from a
% process such as the Sequentially Markov
% Coalescent~\citep{mcvean2005approximating,marjoram2006fast}
% like ARGweaver~\citep{rasmussen2014genome}, it is true that
% we do obtain a fully resolved ARG with precise
% information about recombination, but it is reasonable
% to question how much information from the data is used
% to support older recombinations, in particular.

% % This idea of a continuum of information about recombination
% % being inferrable and representable is consistent with the ideas
% % discussed by~\cite{shipilina2023origin}. There a haplotype block
% % is synonymous with coalescent nodes in the ARG, which we can generalise
% % as nodes in a gARG with at least two children. In principle, these
% % nodes are identifiable by the mutations that occur on the outbound
% % edges of the node.

% It is helpful to consider the difference between fully
% resolved binary trees and those that contain polytomies.
% The situation is in fact directly analogous: an ARG that
% contains fully resolved recombination information must have
% at two most parents per node.

% In general, discussions tend to contrast an ARG
% on one hand as holding \emph{complete} information about recombination
% (and therefore between-tree correlation structure)
% and a sequence of local trees on the other (with, by implication,
% no direct information about between-tree correlations).
% This is a false dichotomy:
% a \emph{continuum} of information about recombination and between-tree
% correlation structure can be represented and inferred, and all
% such levels of information are useful, even if only for computational
% efficiency.

>>>>

The graph visualisations in Fig.~\ref{fig-simplification} have
three novel features which require some explanation.
Firstly, edge weights (the thickness of the lines joining
nodes) correspond to the length of the inheritance intervals
they are annotated with. This allows us to distinguish
edges that persist across many local trees from those that are
less influential (contrast the edge
$(\noderef{g}, \noderef{h})$
with $(\noderef{g}, \noderef{i})$
in Fig.~\ref{fig-simplification}A).
% FIXME this needs some work
Secondly, node colours denote the number of parents that they
have in the graph, allowing us to easily see roots (those
with zero parents), recombinants (those with two parents)
and more complex situations arising from simplification (see below).
Thirdly, the shading intensity of a node denotes the ``coalescent span'', the
fraction of the node's span (the length of genome in which it
is reachable from the samples in the local trees)
over which it is coalescent (has more than one child). Nodes which
are never locally unary therefore have a coalescent span of 100\%,
whereas nodes in which ancestral material never coalesces have a
coalescent span of 0\%.
% JK: Let's not get into discussions of identifiability or inferability
% here, we're just trying to show the things that exist in the true history.
% (such nodes will be extremely difficult to
% identify from the distribution of shared mutations in the samples).
% This is useful why? I can't come up with a quick way of saying
% this. In some senses it's a measure of identifiability?
% JK: Hmm, not sure we want to jump into inferability here?
% All these features will impact what kind of an ARG we can
% hope to infer from sampled genomes.

Returning to the main topic of this section,
Fig.~\ref{fig-simplification}A is the direct output of a Wright-Fisher
simulation in which we retain all nodes involved in recombination
or common ancestry events. This is the true history, and contains
a very high level of detail, some of which may be considered
redundant (or, from another perspective, unobservable).
In Fig.~\ref{fig-simplification}A the local trees (right)
contain many unary nodes, fewer as  we successively
simplify (Fig.~\ref{fig-simplification}B,C),
until we reach Fig.~\ref{fig-simplification}D, where are none.

The first level of simplification that we can perform is based only
on the graph topology.
An example of graph topology that we may consider redundant
(or non-identifiable) is a
``diamond''~\citep{rasmussen2014genome}
in which the two parent nodes of a recombination immediately
join again into a common ancestor (e.g.~\noderef{j}, \noderef{l}, \noderef{m}
and \noderef{n} in Fig.~\ref{fig-simplification}A).
Unless we are specifically
interested in the recombination event or these ancestral genomes,
the diamond can be replaced by a single edge without loss of
information.
More generally, any subgraph that is singly-connected in both the leafward and
rootward direction (a ``super-diamond'') can be replaced by one edge.
This definition includes the case
of a node that has one inbound and one outbound edge, such as
nodes \noderef{f} and \noderef{h}.
Fig.~\ref{fig-simplification}B shows the result of this type of
graph topology simplification.

Simplifying away diamonds will remove many unary nodes from the
local trees, but there can still be nodes that are unary in all
of the local trees. In particular, a node can represent a recombinant
with multiple parents in the graph but only a single child (e.g.\ node \noderef{n}
in Fig.~\ref{fig-simplification}B), or can represent a common ancestor with
multiple children in the graph but in which no coalescence takes place
in the local trees
(node \noderef{r} in Fig.~\ref{fig-simplification}B).
The distinction between the ``common ancestry'' of two or more genomes
in an ancestral genome and the ``coalescence'' which may or may
not occur in the local trees is
important~\citep{hudson1983testing,kelleher2016efficient}.
Consider \noderef{e} in Fig.~\ref{fig-simplification}A,
for example. We can see from the graph that it is a common
ancestor of samples \noderef{a} and \noderef{b}, but
it does not correspond to any coalescence in the
local trees to the left of position $44$, and is therefore
unary in these three trees.
Such nodes are not singly connected in the graph, but are nevertheless unary in
all of the local trees.
The operation to remove them
therefore requires knowledge not just of the graph topology but also of the
ancestral material associated with the edges.
As we see in Fig.~\ref{fig-simplification}C,
removal of recombinant nodes can produce graph nodes with
more than two parents (e.g.~node \noderef{e}); and likewise, removal of
common ancestor but non-coalescent nodes can produce graph nodes with
more than two children (e.g.~node \noderef{s}). These represent the
merged
\emph{effects} of multiple evolutionary events in a single node (genome), and the
ARG no longer contains the intermediate genomes representing those events.

@@

The ARGs in Fig.~\ref{fig-simplification} contain different
numbers of local trees ($6$, $5$, $5$ and $4$ respectively for A through
D). When we move from A to B the local trees
for the intervals $[44,61)$ and $[61,87)$ are merged because
the only differences between them are their paths through
nodes \noderef{l} and \noderef{m}. These nodes that participated
in the diamond are removed from the ARG, and we have lost
all information about the corresponding recombination at
position 61. Other nodes (e.g.\ \noderef{o} and \noderef{p})
have also been removed but these represent the \emph{parents}
of recombinants. The recombinant nodes themselves
(e.g. \noderef{n}) are still present, and represent precise
information about the time, genomic location and
lineages involved in the recombination event.

@@
